---
title: Filtering Methods In R
author: Aaron Lim Jin Ming
date: '2020-11-04'
slug: filtering-methods-in-r
categories:
  - R
tags:
  - R
description: 'Evaluating the different methods of filtering or subsetting data in R.'
keywords:
  - R
  - Filter
  - Subset
  - Select
stylesheet: post.css
---

One of the *greatest* and *worst* things about using R is that there are so many different methods to tackle the same problem. Every time I become aware of an alternative method, I start to think thoughts such as "Is this better than my method?", "Does this always give the same results as my current method?".

Filtering is one scenario when these thoughts surface, there are multiple ways to filter data but what are the actual differences between them? I wrote this post to formally pen down the differences and to resolve these my doubts.

# Filtering

Filtering is the act of choosing a subset of your current data that fits some criteria. In R, this is the act of selecting/discarding certain rows from a dataframe. As far as I am aware, there are three popular methods to filter data.

To illustrate the differences between these three methods, lets create a small dataframe to illustrate their differences.

```{r}
mydata <- data.frame(x1 = 1:10,
                     x2 = c(1:4,NA,NA,7:10),
                     row.names = NULL)
mydata
```

# Subset Operator `[]`

The most basic way to filter would probably be using the subset operator `[]` in R.

```{r}
# give me the rows where x1 is more than 5
mydata[mydata$x1 > 5,]
```

Here is what happens when we use the subset operator to filter on a column with `NA` variables.

```{r}
# give me the data where x2 is less than 3
mydata[mydata$x2 < 3,]
```

Unexpected things happen with the subset operator. It seems like the `NA` values get caught in the filter and are returned even though `NA` is not less than 3. The reason for this behavior is because evaluating `NA` in any logical statement (except `is.na()`) will produce `NA` as an output.

```{r}
NA > 3
NA == "NA"
NA == NA
```

While this may be the quickest way to filter, it breaks down when there are missing values in the filtering variable.

To prevent this behavior, we need additional conditional statements. But this is a hassle to write and read.

```{r}
# give me the data where x2 is less than 3
mydata[mydata$x2 < 3 & !is.na(mydata$x2),]
```

# `subset()`

The `subset()` function is the second way to filter. It is part of base R so I prefer it over the tidyverse `filter()` as I tend to find myself in work where I may not have the luxury to install every package I need. 

```{r}
# give me the rows where x1 is more than 5
subset(mydata,
       mydata$x1 > 5)
```

Lets see how it handles the missing values in x2 when we filter on x2.

```{r}
# give me the data where x2 is less than 3
subset(mydata,
       mydata$x2 < 3)
```

This time it seems to work! `subset()` behaves as expected.

# `filter()` from *dplyr*

Up last is `filter()` from `dplyr`. I personally use this very infrequently because I picked up R sticking to base R. But I do use the other `tidyverse` functions quite frequently as some of them are indeed very powerful. I would use `filter()` when I find myself inside a pipe (`%>%`) chain and I don't wish to awkwardly exit the pipe to `subset()`, just to pipe the data in again.

```{r}
library(dplyr)
# give me the rows where x1 is more than 5
mydata %>% filter(x1 > 5)
```

Lets see how it handles the missing values in x2 when we filter on x2.

```{r}
# give me the data where x2 is less than 3
mydata %>% filter(x2 < 3)
```

It seems like filter works just as well!

# Summary

Avoid using the subset operators `[]` to filter unless you are sure that the filtering variable has no missing values `NA`.`subset()` and dplyr's `filter()` are preferred as they exclude missing values. Choosing between the two would be personal preference or depend on the context.
